---
title: "Divvy-2021-data"
output: github_document
---


```{r}
library(tidyverse) #helps wrangle data
library(lubridate) #helps wrangle date attributes
library(ggplot2) #helps visualize data

library(readr)
```

## COLLECTING DATA

### Reading all the .csv files for first four months of year 2021 :



Divvy_Trips_2021_01 <- read_csv("...//202101-divvy-tripdata.csv")



Divvy_Trips_2021_02 <- read_csv("...//202102-divvy-tripdata.csv")



Divvy_Trips_2021_03 <- read_csv("...//202103-divvy-tripdata.csv")



Divvy_Trips_2021_04 <- read_csv("...//202104-divvy-tripdata.csv")



```{r echo=FALSE}
Divvy_Trips_2021_01 <- read_csv("C:/Users/piyus/Desktop/bike-share PROJECT//202101-divvy-tripdata.csv")



Divvy_Trips_2021_02 <- read_csv("C:/Users/piyus/Desktop/bike-share PROJECT//202102-divvy-tripdata.csv")



Divvy_Trips_2021_03 <- read_csv("C:/Users/piyus/Desktop/bike-share PROJECT//202103-divvy-tripdata.csv")



Divvy_Trips_2021_04 <- read_csv("C:/Users/piyus/Desktop/bike-share PROJECT//202104-divvy-tripdata.csv")


```





## WRANGLE DATA AND COMBINE INTO A SINGLE FILE


### Removing all the unecessary columns (start_lat,start_lng, end_lat, end_lng) :

```{r}
Divvy_Trips_2021_01<- Divvy_Trips_2021_01 %>%
  select(-c(start_lat, start_lng, end_lat, end_lng))


Divvy_Trips_2021_02<- Divvy_Trips_2021_02 %>%
  select(-c(start_lat, start_lng, end_lat, end_lng))


Divvy_Trips_2021_03<- Divvy_Trips_2021_03 %>%
  select(-c(start_lat, start_lng, end_lat, end_lng))


Divvy_Trips_2021_04 <- Divvy_Trips_2021_04 %>%
  select(-c(start_lat, start_lng, end_lat, end_lng))
```


### Changing station_id columns to double format : 


```{r warning=FALSE}
Divvy_Trips_2021_01$start_station_id <- as.double(Divvy_Trips_2021_01$start_station_id)
Divvy_Trips_2021_01$end_station_id <- as.double(Divvy_Trips_2021_01$end_station_id)

Divvy_Trips_2021_02$start_station_id <- as.double(Divvy_Trips_2021_02$start_station_id)
Divvy_Trips_2021_02$end_station_id <- as.double(Divvy_Trips_2021_02$end_station_id)

Divvy_Trips_2021_03$start_station_id <- as.double(Divvy_Trips_2021_03$start_station_id)
Divvy_Trips_2021_03$end_station_id <- as.double(Divvy_Trips_2021_03$end_station_id)

Divvy_Trips_2021_04$start_station_id <- as.double(Divvy_Trips_2021_04$start_station_id)
Divvy_Trips_2021_04$end_station_id <- as.double(Divvy_Trips_2021_04$end_station_id)
```


### Changing the format of started_at and ended_at columns to date_time : 
```{r}
Divvy_Trips_2021_01$started_at <- dmy_hm(Divvy_Trips_2021_01$started_at)
Divvy_Trips_2021_01$ended_at <- dmy_hm(Divvy_Trips_2021_01$ended_at)
View(Divvy_Trips_2021_01)
```

### Creating the ride_length column for all the month files :
```{r}
Divvy_Trips_2021_01$ride_length <- difftime(Divvy_Trips_2021_01$ended_at,Divvy_Trips_2021_01$started_at)
Divvy_Trips_2021_02$ride_length <- difftime(Divvy_Trips_2021_02$ended_at,Divvy_Trips_2021_02$started_at)
Divvy_Trips_2021_03$ride_length <- difftime(Divvy_Trips_2021_03$ended_at,Divvy_Trips_2021_03$started_at)
Divvy_Trips_2021_04$ride_length <- difftime(Divvy_Trips_2021_04$ended_at,Divvy_Trips_2021_04$started_at)
```


### reordering the columns :

```{r}
Divvy_Trips_2021_01 <- Divvy_Trips_2021_01[,c(1,2,3,4,10,5:9)]
Divvy_Trips_2021_02 <- Divvy_Trips_2021_02[,c(1,2,3,4,10,5:9)]
Divvy_Trips_2021_03 <- Divvy_Trips_2021_03[,c(1,2,3,4,10,5:9)]
Divvy_Trips_2021_04 <- Divvy_Trips_2021_04[,c(1,2,3,4,10,5:9)]
```



### adding date,day,month,year columns :

```{r}
Divvy_Trips_2021_01$date <- format(as.Date(Divvy_Trips_2021_01$started_at, "%Y-%m-%d"), "%d-%m-%Y" )
Divvy_Trips_2021_01$month <- format(as.Date(Divvy_Trips_2021_01$date), "%m")
Divvy_Trips_2021_01$day <- format(as.Date(Divvy_Trips_2021_01$date), "%d")
Divvy_Trips_2021_01$year <- format(as.Date(Divvy_Trips_2021_01$date), "%Y")
Divvy_Trips_2021_01$day_of_week <- format(as.Date(Divvy_Trips_2021_01$date), "%A")


Divvy_Trips_2021_02$date <- format(as.Date(Divvy_Trips_2021_02$started_at, "%Y-%m-%d"), "%d-%m-%Y" )
Divvy_Trips_2021_02$month <- format(as.Date(Divvy_Trips_2021_02$date), "%m")
Divvy_Trips_2021_02$day <- format(as.Date(Divvy_Trips_2021_02$date), "%d")
Divvy_Trips_2021_02$year <- format(as.Date(Divvy_Trips_2021_02$date), "%Y")
Divvy_Trips_2021_02$day_of_week <- format(as.Date(Divvy_Trips_2021_02$date), "%A")



Divvy_Trips_2021_03$date <- format(as.Date(Divvy_Trips_2021_03$started_at, "%Y-%m-%d"), "%d-%m-%Y" )
Divvy_Trips_2021_03$month <- format(as.Date(Divvy_Trips_2021_03$date), "%m")
Divvy_Trips_2021_03$day <- format(as.Date(Divvy_Trips_2021_03$date), "%d")
Divvy_Trips_2021_03$year <- format(as.Date(Divvy_Trips_2021_03$date), "%Y")
Divvy_Trips_2021_03$day_of_week <- format(as.Date(Divvy_Trips_2021_03$date), "%A")



Divvy_Trips_2021_04$date <- format(as.Date(Divvy_Trips_2021_04$started_at, "%Y-%m-%d"), "%d-%m-%Y" )
Divvy_Trips_2021_04$month <- format(as.Date(Divvy_Trips_2021_04$date), "%m")
Divvy_Trips_2021_04$day <- format(as.Date(Divvy_Trips_2021_04$date), "%d")
Divvy_Trips_2021_04$year <- format(as.Date(Divvy_Trips_2021_04$date), "%Y")
Divvy_Trips_2021_04$day_of_week <- format(as.Date(Divvy_Trips_2021_04$date), "%A")
```



### Merging all month files into one quarter file of 2021 :

```{r}
Divvy_2021_Q1_data <- bind_rows(Divvy_Trips_2021_01 , Divvy_Trips_2021_02,Divvy_Trips_2021_03,Divvy_Trips_2021_04)
View(Divvy_2021_Q1_data)
```




## CONDUCT DESCRIPTIVE ANALYSIS

### Descriptive analysis on ride_length (all figures in seconds)

```{r}
mean(Divvy_2021_Q1_data$ride_length) 
median(Divvy_2021_Q1_data$ride_length) 
max(Divvy_2021_Q1_data$ride_length)
min(Divvy_2021_Q1_data$ride_length) 
```


### Summary of the ride length column :
```{r}
summary(Divvy_2021_Q1_data$ride_length)
```
### Comparing  members and casual riders :

```{r}
aggregate(Divvy_2021_Q1_data$ride_length ~ Divvy_2021_Q1_data$member_casual, FUN = mean)
aggregate(Divvy_2021_Q1_data$ride_length ~ Divvy_2021_Q1_data$member_casual, FUN = median)
aggregate(Divvy_2021_Q1_data$ride_length ~ Divvy_2021_Q1_data$member_casual, FUN = max)
aggregate(Divvy_2021_Q1_data$ride_length ~ Divvy_2021_Q1_data$member_casual, FUN = min)
```
### Tthe average ride time by each day for members vs casual riders :


### Ordering the days of week :

```{r}
Divvy_2021_Q1_data$day_of_week <- ordered(Divvy_2021_Q1_data$day_of_week, levels=c("Sunday", "Monday",
"Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```
### the average ride time by each day for members vs casual riders :

```{r}
aggregate(Divvy_2021_Q1_data$ride_length ~ Divvy_2021_Q1_data$member_casual + Divvy_2021_Q1_data$day_of_week,
FUN = mean)
```


### Analyzing ridership data by type and weekday :

```{r}
Divvy_2021_Q1_data %>%
mutate(weekday = wday(started_at, label = TRUE)) %>% 
group_by(member_casual, weekday) %>%
summarise(number_of_rides = n(),average_duration = mean(ride_length)) %>%
arrange(member_casual, weekday)
```


###   Visualizing the number of rides by rider type :
```{r}
Divvy_2021_Q1_data %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(),average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(mapping=aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```




### Visualization for average duration :

```{r}
Divvy_2021_Q1_data%>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(),average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")

```




### Comparing general bike type preference between members and casual riders :

```{r}
Divvy_2021_Q1_data%>% 
  group_by(rideable_type, member_casual) %>% 
  summarize(number_of_rides = n(), .groups = 'drop')

```



### Comparing number of docked_bike rides between members and casual riders for each day of week
```{r}
Divvy_2021_Q1_data%>%
  filter(rideable_type == 'docked_bike') %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(number_of_rides = n(), .groups = 'drop') %>% 
  arrange(day_of_week)
```



### Docked_bike rides of casual riders for each day of week :

```{r}
Divvy_2021_Q1_data %>% 
  filter(rideable_type == 'docked_bike', member_casual == 'casual') %>% 
  group_by(day_of_week) %>% 
  summarise(number_of_rides = n(), .groups = 'drop')

```

### Creating a custom theme :

```{r}
custom_theme = theme(plot.title=element_text(size=20),
                 axis.text.x=element_text(size=10), 
                 axis.text.y=element_text(size=12),
                 axis.title.x=element_text(size=18), 
                 axis.title.y=element_text(size=18),
                 strip.text.x=element_text(size=16), 
                 strip.text.y=element_text(size=16),
                 legend.title=element_text(size=18), 
                 legend.text=element_text(size=16))

```



### Average Number of Rides by Month: Casual Riders

```{r}
Divvy_2021_Q1_data%>% 
  group_by(month, member_casual) %>% 
  summarize(number_of_rides = n(), .groups = 'drop') %>% 
  filter(member_casual == 'casual') %>%
  drop_na() %>%
  ggplot(aes(x = month, y = number_of_rides,fill=member_casual)) + 
  geom_bar(position = 'dodge', stat = 'identity') + scale_y_continuous(labels = scales::comma) +
  theme(axis.text.x = element_text(angle = 45)) + 
  labs(x = "Month", y = "Number of Rides", 
       fill = "Member/Casual",
       title = "Average Number of Rides by Month: Casual Riders") +
  custom_theme
```






###  average number of rides by hour (casual riders) :

```{r}
options(repr.plot.width = 12, repr.plot.height = 8)

Divvy_2021_Q1_data %>%
  filter(member_casual == 'casual') %>%
  mutate(hour_of_day = hour(round_date(started_at, 'hour'))) %>% 
  group_by(hour_of_day, member_casual) %>% 
  summarize(number_of_rides = n(), .groups = 'drop') %>% 
  arrange(-number_of_rides) %>% 
  ggplot(aes(x = hour_of_day, y = number_of_rides, fill = member_casual)) +
  geom_bar(position = 'dodge', stat = 'identity',color="black") + scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = c(0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23)) +
  labs(x = "Time of the Day (h)", y = "Number of Rides", fill = "Member/Casual",
       title = "Average Number of Rides by Hour: Casual Riders") +
 custom_theme

```



### Top 10 start stations (members)

```{r}
options(repr.plot.width = 10, repr.plot.height = 6)

Divvy_2021_Q1_data %>% 
  group_by(start_station_name, member_casual) %>% 
  summarize(number_of_rides = n(), .groups = 'drop') %>% 
  filter(start_station_name != "", member_casual != "casual") %>% 
  arrange(-number_of_rides) %>% 
  head(n=10) %>%
  ggplot(aes(x = reorder(start_station_name, number_of_rides), y = number_of_rides)) +
  geom_col(position = 'dodge',fill="darkblue" , color= "lightyellow" ) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = 'Top 10 Start Stations for Members', x = '', y = "Number of Rides") +
  coord_flip() +
  theme_minimal() +
  custom_theme

```




### Top 10 start stations (casual riders) :

```{r}
options(repr.plot.width = 10, repr.plot.height = 6)

Divvy_2021_Q1_data%>% 
  group_by(start_station_name, member_casual) %>% 
  summarize(number_of_rides = n(), .groups = 'drop') %>% 
  filter(start_station_name != "", member_casual != "member") %>% 
  arrange(-number_of_rides) %>% 
  head(n=10) %>%
  ggplot(aes(x = reorder(start_station_name, number_of_rides), y = number_of_rides)) +
  geom_col(position = 'dodge', fill ="darkorange1",color="black") +
  scale_y_continuous(labels = scales::comma) +
  labs(title = 'Top 10 Start Stations for Casual Riders', x = '', y = "Number of Rides") +
  coord_flip() +
  theme_minimal() +
  custom_theme

```



### Usage of different bikes by rider type :

```{r}
options(repr.plot.width = 12, repr.plot.height = 8)

Divvy_2021_Q1_data %>% 
  group_by(rideable_type, member_casual) %>% 
  summarize(number_of_rides = n(), .groups = 'drop') %>% 
  drop_na() %>% 
  ggplot(aes(x = member_casual, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values=c("casual"="gold",  "member"="navyblue"))+
  scale_y_continuous(labels = scales::comma) +
  facet_wrap(~rideable_type) +
  labs(fill = "Member/Casual", x = "", y = "Number of Rides", 
       title = "Usage of Different Bikes: Members vs. Casual Riders") + custom_theme

```



###  Usage of different bikes by rider type (separated) :


```{r}
options(repr.plot.width = 14, repr.plot.height = 10)

Divvy_2021_Q1_data%>% 
  group_by(month, member_casual, rideable_type) %>% 
  summarize(number_of_rides = n(), .groups = 'drop') %>% 
  drop_na() %>% 
  ggplot(aes(x = month, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values=c("casual"="darkred","member"="turquoise4" ))+
  scale_y_continuous(labels = scales::comma) +
  facet_grid(member_casual~rideable_type) +
  labs(x = "Month", y = "Number of Rides", fill = "Member/Casual",
       title = "Average Number of Rides by Month") +
  theme(axis.text.x = element_text(angle = 90)) + custom_theme

```



###  Number of rides between members and casual riders by day of week across the year:

```{r}
options(repr.plot.width = 26, repr.plot.height = 10)

Divvy_2021_Q1_data %>% 
  mutate(weekday=wday(started_at, label = TRUE))%>%
  group_by(month, weekday, member_casual) %>% 
  summarize(number_of_rides = n(), .groups = 'drop') %>% 
  drop_na() %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values=c("casual"= "tomato","member"="turquoise4"))+
  scale_y_continuous(labels = scales::comma) +
  facet_grid(member_casual~month) +
  labs(x = "Day of Week", y = "Number of Rides", fill = "Member/Casual",
       title = "Bike Usage between Members and Casual Riders by Day of Week across the Year", fill = 'Member/Casual') +
  theme(axis.text.x = element_text(angle = 90)) +
  custom_theme

```




























